---
title: "PathAnalyser"
author:
  - name: Rishab Kaushik
  - name: Taniya Pal
  - name: Ozlem Karadeniz
  - name: Yi-Hsuan Lee
  - name: Anisha Thind
  - name: Justice Iheanyichi
abstract: >
  PathAnalyser provides functionality for assessing ER and HER2 pathway activity
  in breast cancer transcriptomic datasets. The package enables classification 
  of samples in transciptomic datasets according to pathway activity by using a 
  gene expression signature associated with a pathway, a list of genes, which 
  vary in expression depending on the pathway activity. These transcriptional 
  signatures have been shown to be of clinical significance, for example, ER and 
  HER2 gene signatures have been associated with molecular sub-types, prognosis 
  and treatment reponse in breast cancer. Although this package was original 
  developed to assess ER and HER2 pathway activity in breast cancer 
  transcriptomic datasets, the package functionality could be applied to 
  transcriptomic datasets and gene signatures outside the context of breast 
  cancer. In this vignette, we will describe how to use the PathAnalyser package 
  with microarray and RNA-seq expression transcriptomic datasets and gene 
  expression signatures associated with a specific pathway activity.
output: 
  BiocStyle::html_document:
    toc: TRUE
    toc_float: TRUE
   
vignette: >
  %\VignetteIndexEntry{PathAnalyser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```
# Installation and setup
To install a local source version of the package, type the following R:
```{r eval=FALSE}
install.packages("path-to-source/PathAnalyser-main.zip", repo=NULL, type="source")
```

Alternatively, the package can be installed directly from GitHub using the
devtools package in R:
```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("a-thind/PathAnalyser")
```

Once installation is completed, load the library to start using PathAnalyser:
```{r setup}
library(PathAnalyser)
```

# Overview of PathAnalyser functionality
The PathAnalyser package functionality can be sub-divided into 5 broad 
categories corresponding to a typical workflow for assessing pathway activity of 
samples:

1.  Input functions for expression data and gene signatures
2.  Quality control and pre-processing of input data
3.  Classification of samples by pathway activity
4.  Evaluation of classification
5.  Visualization

# Built-in datasets and gene 
## Gene expression signatures
PathAnalyser contains two gene signatures for active ER and HER2 pathways that 
can be used to assess ER and HER2 pathway activity in a breast cancer 
transcriptomic data set. These signatures are accessible by typing the following 
in R:
```{r}
data("ER_sig_df")
data("HER2_sig_df")
```
### ER signature
The built-in ER signature `ER_sig_df` is a data frame containing the names of 
160 differentially expressed genes which constitute the transcriptional 
signature of ER pathway activation. This signature was obtained from the 
sensitivity to endocrine therapy genome index defined by Symanns et al. (2011). 
The total number of genes that are up-regulated (denoted with an expression 
value of 1) and down-regulated (denoted with an expression value of -1) in this 
gene signature are 101 and 59 genes respectively.
```{r}
dim(ER_sig_df)
head(ER_sig_df)
# up-regulated genes are given an expression value of 1
ER_up_sig <- ER_sig_df[ER_sig_df$expression == 1 ,]
dim(ER_up_sig)
# down-regulated genes are given an expression value of -1
ER_dn_sig <- ER_sig_df[ER_sig_df$expression == -1 ,]
dim(ER_dn_sig)
# for more details
?ER_sig_df
```

### HER2 signature
A gene expression signature for HER2 (ERBB2) pathway activation `HER2_sig_df` is 
also provided by PathAnalyser. This gene signature was obtained from Molecular 
Signatures Database ([MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp)) 
by combining the up-regulated (gene set:
[ERBB2_UP.V1_UP](https://www.gsea-msigdb.org/gsea/msigdb/cards/ERBB2_UP.V1_UP.html) 
) and down-regulated (gene set:
[ERBB2_UP.V1_DN](https://www.gsea-msigdb.org/gsea/msigdb/geneset_page.jsp?geneSetName=ERBB2_UP.V1_DN) 
) gene sets available at MSigDB, originating from transcriptomic profiling of a 
MCF-7 breast cancer cell line, positive for the estrogen receptor (ER) but 
genetically engineered to express the ERBB2 (HER2) receptor. 

The names of 387 differentially expressed genes are in the HER2 signature data 
frame provided by PathAnalyser, of which 190 are up-regulated 
(denoted by an expression of 1) and 197 are down-regulated (denoted by an 
expression of -1).
```{r}
dim(HER2_sig_df)
head(HER2_sig_df)
# up-regulated genes are given an expression value of 1
HER2_up_sig <- HER2_sig_df[HER2_sig_df$expression == 1 ,]
dim(HER2_up_sig)
# Down-regulated genes are given an expression value of -1
HER2_dn_sig <- HER2_sig_df[HER2_sig_df$expression == -1 ,]
dim(HER2_dn_sig)
# for more details
?HER2_sig_df
```
## Gene expression datasets
PathAnalyser also contains two built-in gene expression matrices each containing
RNA-seq raw read counts for primary breast cancer samples obtained from 20 
individuals (cases). Data for these matrices were obtained from The Cancer 
Genome Atlas ([TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga)).
Each expression matrix contains 20,124 genes.
```{r}
data("ER_data_se1")
data("HER2_data_se1")
# column names represent case (sample) IDs from TCGA
```

### ER dataset 1

```{r}
dim(ER_data_se1)
# Expression data for first 6 genes
head(ER_data_se1)
```

### HER2 dataset 1
Need to get info
```{r}
dim(HER2_data_se1)
# Expression data for first 6 genes
head(HER2_data_se1)
```


# Reading input data using PathAnalyser
There are two types of input required for pathway activity assessment using 
PathAnalyser:

1.  gene expression data matrices
2.  gene expression signatures

## Gene expression data
The `read_expression_data` function reads gene expression data from a 
tab-delimited text file with gene names as the first column and successive 
columns containing gene expression data for each sample in the given data set. 
```{r eval=FALSE}
data_se <- read_expression_data("~/PathAnalyser/raw_data/Toy_dataset.txt")

```

## Gene signature data
Gene signature files that follow the gene set format (ending with .grp) can be 
read by the `read_signature_data` function.
```{r eval=FALSE}
sig_df <- read_signature_data("~/PathAnalyser/raw_data/ESR1_UP.v1._UP.grp", "~/PathAnalyser/raw_data/ESR1_DN.v1_DN.grp")
```

# Quality control and pre-processing of data
For the remainder of this demonstration, we will use the built-in gene 
expression data set matrix `ER_data_se1`.
```{r}
data_se <- ER_data_se1 # stand-in until input functions are completed
evaluatematrix(data_se) # need to change name to evaluate_matrix
normalized_se <- transform(data_se)
# Decide on whether the singature data frame needs to be split before or inside
# function
#data_se <- check_signature_vs_dataset(normalized_se, as.data.frame(up),as.data.frame(dn))
```

# Classifying samples by pathway activity using gene signatures
## Visualising GSVA Score Distribution
To visualise the distribution of GSVA scores for the up-regulated gene signature
and the down-regulated gene signature, use the `visualise_GSVA` function:
```{r}
visualise_GSVA(ER_sig_df, data_se)
```


## Classification using absolute GSVA score thresholds
```{r}
classes_df.abs <- classify_GSVA_abs(ER_sig_df, data_se, up_thresh.low=-0.25,
                                    up_thresh.high=0.25, dn_thresh.low=-0.25,
                                    dn_thresh.high=0.25)
```

## Classification using percentile GSVA score thresholds
```{r}
classes_df.percent <- classify_GSVA_percent(ER_sig_df, data_se)
```

# Evaluating pathway activity classification
```{r}
#confusion_matrix_HER2 <- calculate_accuracy(labelDF, classes_df.abs)
```
# Visualising pathway activity
## Generating PCA plot
```{r}
classes_pca(normalized_se, classes_df.abs, pathway_name = "ER")
```

# Session Info
```{r}
sessionInfo()
```

# References
Hänzelmann, S., Castelo, R. & Guinney, J. GSVA: gene set variation analysis for 
microarray and RNA-Seq data. BMC Bioinformatics 14, 7 (2013). 
https://doi.org/10.1186/1471-2105-14-7

Symmans, W.F., Hatzis, C., Sotiriou, C., Andre, F., Peintinger, F., Regitnig, 
P., Daxenbichler, G., Desmedt, C., Domont, J., Marth, C. and Delaloge, S., 
2010. Genomic index of sensitivity to endocrine therapy for breast cancer. 
Journal of clinical oncology, 28(27), p.4111. 
https://dx.doi.org/10.1200%2FJCO.2010.28.4273
